\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \footnote{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-10-10 move_schedules.tex
Edits to streamline section 3. Added reviewer point-by-point responses to most of the reviewer feedbakc so far

% --- Springer only ---
\author[1]{\fnm{Matthew} \sur{Piekenbrock}}\email{piekenbrock.m@northeastern.edu}
\affil[1]{
	\orgdiv{Khoury College of Computer Sciences}, 
	\orgname{Northeastern University}
}

% --- Springer only --- 
\author[2]{\fnm{Jose} \sur{A. Perea}}\email{j.pereabenitez@northeastern.edu}
\affil[2]{
	\orgdiv{Department of Mathematics and Khoury College of Computer Sciences},
	\orgname{Northeastern University}
	% \orgaddress{\street{360 Huntington Ave}, \city{Boston}, \postcode{02115}, \state{MA}}
}

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract} % arxiv only 
\abstract{ % springer only
Matrix reduction is the standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computing persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations. By reduction to a particular longest common subsequence problem, we show that the minimal number $d$ of updates to the decomposition can be found in $O(m \log \log m)$ time and $O(m)$ space, and that the corresponding sequence of permutations---which we call a \emph{schedule}---can be constructed in $O(d m \log m)$ time.  
We also show that, in expectation, the storage needed to employ this strategy is actually sublinear in $m$. 
Exploiting this connection, we show experimentally that the decrease in operations to compute diagrams across a family of filtrations, is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multi-parameter persistence are also presented.
} % springer only
%\end{abstract} % arxiv only

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With state of the art tools, there is no recourse when faced with a time-varying complex containing millions of simplices across thousands of snapshots in time.
Moreover, acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes all immediately become computationally viable tools in dynamic settings. Such persistence summaries include adaptive template functions~\cite{polanco2019adaptive}, persistence images~\cite{adams2017persistence}, and $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagrams of a filtered simplicial complex $K$ with $m$ simplices in $O(m^3)$ time, via a particular decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm, in turn,  transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, the initial reduction cost is amortized by the cost of maintaining the decomposition, implying each diagram produced by vineyards takes just linear time per time point. 
% approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc. see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 does have an implementation of vineyards, however the algorithm was never ported to version 2. Other major packages, such as GUDHI and PHAT, do not have vineyards implementations.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, they observe that maintaining the decomposition along a certain parameterized family is the most computationally demanding aspect of RIVET~\cite{rivet}, a software for computing and visualizing	 two-parameter persistent homology.

The work presented here seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
Though the vineyards algorithm is efficient at constructing a \emph{continuous} 1-parameter family of diagrams, it is not necessarily efficient when the parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ For all accompanying software and materials, see: \url{https://github.com/peekxc/move_schedules}}

@@

If  zeroth homology  is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the edge are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give an algorithm that maintains a \emph{merge tree} with $e$ edges in $O(e)$ time per-update.
If only Betti numbers are needed, the zeroth-dimension problem reduces even further to the \emph{dynamic connectivity problem}, which can be be efficiently solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

@@

We use this data set as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
Since it is unknown a priori when the persistent pairing function changes, one solution is to do $n$ independent persistence computations at $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between a pair of filtrations $(K, f)$, $(K,f')$ and then decompose this homotopy into adjacent transpositions based on the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, and the  latter as the \emph{continuous setting}. Note that though the discrete setting is often more practical, it is not guaranteed to capture all homological changes in persistence that occur in the continuous 1-parameter family of diagrams.

@@

Note that without knowing where the persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than vineyards, while still capturing the main events.

The existence of a time discretization that is more efficient to compute than continually updating the decomposition, indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when the pairing function determining the persistence diagram is unchanged. 
Indeed, as shown by the case where $n = 10$, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is substantially more efficient than iteratively updating the decomposition.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to vineyards is clearly needed. 
 \\
 \\
 \noindent
\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations for dynamic persistence settings. 
By a reduction to an edit distance problem, we give a   lower bound on the minimal number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition,  along with a proof of its optimality.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for constructing these operations---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
These operations parameterize sequences of permutations $\mathcal{S}= ( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for heuristically solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

We now summarize our main results  (Theorem~\ref{prop:sparsity_result}): suppose one has a  schedule  $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where  $s_k = (i_k, j_k)$ for $k=1,\ldots, d$,  denotes a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_k < j_k$ for all $s_k \in \mathcal{S}$, our first result extends~\cite{busaryev2010tracking} by showing that~\eqref{eq:rv_seq} can be computed  using $O(\kappa)$ column operations, where: 
\begin{equation}
	\quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert 
\end{equation}
The quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $f$ and $f'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself and is thus output sensitive, which we leverage in Section~\ref{sec:schedule_cost}. 
%Surprisingly, due to a refined analysis of restore subprocedure from~\cite{busaryev2010tracking} we give in section~\ref{sec:continuous_setting}, we find that move operations are actually cheaper than vineyards, see Proposition~\ref{prop:factor2}.
%if the vineyards algorithm requires $T$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that obtains the same sequence using at most $M$ column operations, where: 
%\begin{equation}
%	M \leq \tfrac{1}{2}  T
%\end{equation}

@@

\subsection{The Reduction Algorithm}\label{sec:reduction}
In this section we briefly recount the original reduction algorithm introduced in~\cite{zomorodian2005computing}, also sometimes called the \emph{standard} algorithm or more explicitly \texttt{pHcol}~\cite{de2011dualities}. 
The pseudocode is outlined in Algorithm~\ref{alg:reduce} in the appendix. Without optimizations, like   clearing   or implicit matrix reduction, the standard  algorithm is very inefficient. Nonetheless, it serves as the foundation of most persistent homology implementations, and its invariants are necessary before introducing both vineyards in section~\ref{sec:vineyards} and our move schedules  in section~\ref{sec:move_schedules}.

@@

Cohen-Steiner et al.~\cite{cohen2007stability} referred to these curves as \emph{vines}, a collection of which forms as \emph{vineyard}---the geometric analogy meant to act as a guidepost for practitioners seeking to understand the evolution of topological structure over time.

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, detecting homological critical events along the way.
As homological critical events only occur when the filtration order changes, detecting all such events may be reduced to computing valid decompositions at   time points interleaving all changes in the filtration order. 
For simplexwise filtrations, these changes manifest as transpositions of adjacent simplices, and thus any fixed set of rules that maintains a valid $R = D V$ decomposition under adjacent column transpositions is sufficient to compute persistence dynamically.

@@

\textbf{Example 2.2} To illustrate the basic principles of vineyards, we re-use the running example introduced in the previous section. Below, we illustrate the case of exchanging simplices $a$ and $b$ in the filtration order, and restoring $RV$ to a valid decomposition. 

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition.
Formally, since column operations are the largest complexity operations needed and each column can have potentially $O(m)$ entries, the complexity of vineyards is $O(m)$ per transposition. 
Inspection of the individual cases of the algorithm from~\cite{cohen2006vines} shows that any single transposition requires at most two such operations on both $R$ and $V$.
However, several factors can affect the runtime efficiency of the vineyards algorithm. 
On the positive side, as both the $V$ and $R$ matrices are often sparse, the cost of a given column operation is proportional to the number of non-zero  entries in the two columns being modified. Moreover, as a rule of thumb, its been observed that most transpositions require no column operations~\cite{edelsbrunner2000topological}. 
On the negative side, one needs to frequently query the non-zero status of various entries in $R$ and $V$ (consider evaluating e.g. Case 1.1 in~\cite{cohen2006vines}), which accrues a non-trivial runtime cost due to the quadratic frequency with which they are required.   
% Moreover, an efficient implementation of the vineyards algorithm requires a non-standard sparse matrix representation that allows swapping any two rows and columns in $O(1)$, as described in Figure 5 in~\cite{cohen2006vines}. 
% Though these minor operations do not affect the asymptotic complexity of the vineyards algorithm, they collectively accrue a non-trivial runtime cost

@@

\subsection{Moves}\label{sec:moves} Originally developed to accelerate tracking generators with temporal coherence, Busaryev et al.~\cite{busaryev2010tracking} introduced an extension of the vineyards algorithm which maintains a $R = D V$ decomposition under \emph{move operations}. A move operation $\mathrm{Move}(i,j)$ is a set of rules for maintaining a valid decomposition under the permutation $P$ that moves a simplex $\sigma_i$ at position $i$ to position $j$. If $j = i \pm 1$, this operation is an adjacent transposition, and in this sense moves generalizes vineyards.
However, the move framework presented by Busaryev is actually distinct in that it exhibits several attractive qualities not inherited by the vineyards approach that warrants further study.

@@

Similarly, in the context of edit operations, observe that a move operation can be interpreted as a paired delete-and-insert operation, i.e. $m_{ij} = (\mathrm{ins}_j \circ \mathrm{del}_i)$, where $\mathrm{del}_i$ denotes the operation that deletes the character at position $i$  and $\mathrm{ins}_j$ the operation that inserts the same character at position $j$. 

@@

Moreover, $\lvert S^\ast \rvert = d$ can be determined in $O(m \log \log m)$ time. 

>>>>

% --- Springer only ---
\author[1]{\fnm{Matthew} \sur{Piekenbrock}}\email{piekenbrock.m@northeastern.edu}
\affil[1]{
	\orgdiv{Khoury College of Computer Sciences}, 
	\orgname{Northeastern University}}

% --- Springer only --- 
\author[2]{\fnm{Jose} \sur{A. Perea}}\email{j.pereabenitez@northeastern.edu}
\affil[2]{
	\orgdiv{Department of Mathematics and Khoury College of Computer Sciences},
	\orgname{Northeastern University}}
% \orgaddress{\street{360 Huntington Ave}, \city{Boston}, \postcode{02115}, \state{MA}}
%}

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract} % arxiv only 
\abstract{ % springer only
Matrix reduction is the standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computing persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations. By reduction to a particular longest common subsequence problem, we show that the minimal number of decomposition updates $d$ can be found in $O(m \log \log m)$ time and $O(m)$ space, and that the corresponding sequence of permutations---which we call a \emph{schedule}---can be constructed in $O(d m \log m)$ time.  
We also show that, in expectation, the storage needed to employ this strategy is actually sublinear in $m$. 
Exploiting this connection, we show experimentally that the decrease in operations to compute diagrams across a family of filtrations, is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multi-parameter persistence are also presented.
} % springer only
%\end{abstract} % arxiv only

@@

\section{Introduction} 
%\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
Currently, there is no recourse when faced with a time-varying complex containing millions of simplices across thousands of snapshots in time.
%---the state of the art is just to compute the persistence at each time point independently. % ask jose what he thinks
Acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes all immediately become computationally viable tools in the dynamic setting. Such persistence summaries include adaptive template functions~\cite{polanco2019adaptive}, persistence images~\cite{adams2017persistence}, and $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagrams of a filtered simplicial complex $K$ with $m$ simplices in $O(m^3)$ time, via a particular decomposition $R = D V$ (or $RU = D$) of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm, in turn,  transforms a time-varying filtration into a certain set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
Moreover, the vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  \emph{vineyards}  takes $O(m^3 + md)$ time; one initial $O(m^3)$-time reduction at time $t_0$ followed by one $O(m)$ operation to update the decomposition at the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, the initial reduction cost is amortized by the cost of maintaining the decomposition, implying each diagram produced takes just linear time per time point to obtain. 
% approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams at every time point independently.

Despite its theoretical efficiency, \emph{vineyards} is often not the method of choice in practical settings. 
While there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc. see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 does have an implementation of \emph{vineyards}, however the algorithm was never ported to version 2. Other major packages, such as GUDHI and PHAT, do not have \emph{vineyards} implementations.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, they observe that maintaining the decomposition along a certain parameterized family is the most computationally demanding aspect of RIVET~\cite{rivet}, a software for computing and visualizing	 two-parameter persistent homology.

The work presented here seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
Though the vineyards algorithm is efficient at constructing a \emph{continuous} 1-parameter family of diagrams, it is not necessarily efficient when the parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than \emph{vineyards} in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ For all accompanying software and materials, see: \url{https://github.com/peekxc/move_schedules}}

@@

If  zeroth homology  is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the edges of the graph are in filtered order a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give an algorithm that maintains a \emph{merge tree} with $e$ edges in $O(e)$ time per-update.
If only Betti numbers are needed, the zeroth-dimension problem reduces even further to the \emph{dynamic connectivity problem}, which can be be efficiently solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

@@

We use this data set as a baseline for comparing \emph{vineyards} and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
Since it is unknown a priori when the persistent pairing function changes, one solution is to do $n$ independent persistence computations at $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between a pair of filtrations $(K, f)$, $(K,f')$ and then decompose this homotopy into adjacent transpositions based on the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, and the  latter as the \emph{continuous setting}. Note that though the discrete setting is often more practical, it is not guaranteed to capture all homological changes in persistence that occur in the continuous 1-parameter family of diagrams.

@@

Note that without knowing where the persistence pairing function changes, a continuous strategy must construct all $\approx 7 \times 10^{4}$ diagrams induced by the homotopy. 
In this setting, as shown in the figure, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than \emph{vineyards}, while still capturing the main events.

The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards framework must incur some overhead (in terms of column operations) to maintain the underlying decomposition, even when the pairing function determining the persistence diagram is unchanged. 
Indeed, as shown by the case where $n = 10$, applying \texttt{pHcol} independently between  relatively ``close'' filtrations is substantially more efficient than iteratively updating the decomposition.
 Moreover, any optimizations to the reduction algorithm (e.g. clearing~\cite{chen2011persistent}) would only increase this disparity.
Since persistence has found many applications in dynamic contexts~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}, a more efficient alternative to \emph{vineyards} is clearly needed. 
 \\
 \\
 \noindent
\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et al.~\cite{busaryev2010tracking} to include  coarser  operations for dynamic persistence settings. 
By a reduction to an edit distance problem, we give a   lower bound on the minimal number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition,  along with a proof of its optimality.
We also give worst-case sizes of these quantities in expectation as well as efficient algorithms for constructing these operations---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
These operations parameterize sequences of permutations $\mathcal{S}= ( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all minimal size schedules incur the same  cost 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d^2 \log m)$ algorithm for heuristically solving this proxy minimization. 
A performance comparison with other reduction-based persistence computations is given, wherein move schedules are demonstrated to be an order of magnitude more efficient than existing approaches at calculating persistence in dynamic settings. In particular, we illustrate the effectiveness of efficient scheduling with a variety of real-world applications, including flock analysis in dynamic metric spaces and manifold detection from image data using 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

We now summarize our main results  (Theorem~\ref{prop:sparsity_result}): suppose one has a  schedule  $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where  $s_k = (i_k, j_k)$ for $k=1,\ldots, d$,  denotes a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_k < j_k$ for all $s_k \in \mathcal{S}$, our first result extends~\cite{busaryev2010tracking} by showing that~\eqref{eq:rv_seq} can be computed  using $O(\nu)$ column operations, where: 
\begin{equation}
	\quad \nu = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert 
\end{equation}
The quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $f$ and $f'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself and is thus output sensitive, which we leverage in Section~\ref{sec:schedule_cost}. 
%Surprisingly, due to a refined analysis of restore subprocedure from~\cite{busaryev2010tracking} we give in section~\ref{sec:continuous_setting}, we find that move operations are actually cheaper than vineyards, see Proposition~\ref{prop:factor2}.
%if the vineyards algorithm requires $T$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that obtains the same sequence using at most $M$ column operations, where: 
%\begin{equation}
%	M \leq \tfrac{1}{2}  T
%\end{equation}

@@

\subsection{The Reduction Algorithm}\label{sec:reduction}
In this section we briefly recount the original reduction algorithm introduced in~\cite{zomorodian2005computing}, also sometimes called the \emph{standard} algorithm or more explicitly \texttt{pHcol}~\cite{de2011dualities}. 
The pseudocode is outlined in Algorithm~\ref{alg:reduce} in the appendix. Without optimizations, like   clearing   or implicit matrix reduction, the standard  algorithm is very inefficient. Nonetheless, it serves as the foundation of most persistent homology implementations, and its invariants are necessary before introducing both \emph{vineyards} in section~\ref{sec:vineyards} and our move schedules in section~\ref{sec:move_schedules}.

@@

Cohen-Steiner et al.~\cite{cohen2007stability} referred to these curves as \emph{vines}, a collection of which forms as \emph{vineyard}---the geometric analogy is meant to act as a guidepost for practitioners seeking to understand the evolution of topological structure over time.

The original purpose of \emph{vineyards}, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, detecting homological critical events along the way.
As homological critical events only occur when the filtration order changes, detecting all such events may be reduced to computing valid decompositions at   time points interleaving all changes in the filtration order. 
For simplexwise filtrations, these changes manifest as transpositions of adjacent simplices, and thus any fixed set of rules that maintains a valid $R = D V$ decomposition under adjacent column transpositions is sufficient to compute persistence dynamically.

@@

\textbf{Example 2.2} To illustrate the basic principles of \emph{vineyards}, we re-use the running example introduced in the previous section. Below, we illustrate the case of exchanging simplices $a$ and $b$ in the filtration order, and restoring $RV$ to a valid decomposition. 

@@

 The time complexity of \emph{vineyards} is determined entirely by the complexity of performing a single adjacent transposition.
Formally, since column operations are the largest complexity operations needed and each column can have potentially $O(m)$ entries, the complexity of \emph{vineyards} is $O(m)$ per transposition. 
Inspection of the individual cases of the algorithm from~\cite{cohen2006vines} shows that any single transposition requires at most two such operations on both $R$ and $V$.
However, several factors can affect the runtime efficiency of the vineyards algorithm. 
On the positive side, as both the $V$ and $R$ matrices are often sparse, the cost of a given column operation is proportional to the number of non-zero  entries in the two columns being modified. Moreover, as a rule of thumb, its been observed that most transpositions require no column operations~\cite{edelsbrunner2000topological}. 
On the negative side, one needs to frequently query the non-zero status of various entries in $R$ and $V$ (consider evaluating e.g. Case 1.1 in~\cite{cohen2006vines}), which accrues a non-trivial runtime cost due to the quadratic frequency with which they are required.   
% Moreover, an efficient implementation of the vineyards algorithm requires a non-standard sparse matrix representation that allows swapping any two rows and columns in $O(1)$, as described in Figure 5 in~\cite{cohen2006vines}. 
% Though these minor operations do not affect the asymptotic complexity of the vineyards algorithm, they collectively accrue a non-trivial runtime cost

@@

\subsection{Moves}\label{sec:moves} Originally developed to accelerate tracking generators with temporal coherence, Busaryev et al.~\cite{busaryev2010tracking} introduced an extension of the vineyards algorithm which maintains a $R = D V$ decomposition under \emph{move operations}. A move operation $\mathrm{Move}(i,j)$ is a set of rules for maintaining a valid decomposition under the permutation $P$ that moves a simplex $\sigma_i$ at position $i$ to position $j$. If $j = i \pm 1$, this operation is an adjacent transposition, and in this sense \emph{moves} generalizes \emph{vineyards}.
However, the move framework presented by Busaryev is actually distinct in that it exhibits several attractive qualities not inherited by the \emph{vineyards} approach that warrants further study.

@@

Observe that a move operation can be interpreted as a paired delete-and-insert operation, i.e. $m_{ij} = (\mathrm{ins}_j \circ \mathrm{del}_i)$, where $\mathrm{del}_i$ denotes the operation that deletes the character at position $i$  and $\mathrm{ins}_j$ the operation that inserts the same character at position $j$. 

@@
